Communicative reduction in referring expressions within a multi-player negotiation game

### 2023-04-12 write-ups/cogsci23_paper/manuscript.Rmd
cogsci final

output: cogsci2016::cogsci_paper
#final-submission: \cogscifinalcopy
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=3, fig.height=3, fig.crop = F, 
                      fig.pos = "tb", fig.path='figs/',
                      echo=F, warning=F, cache=F, 
                      message=F, sanitize = T)
library(here)
library(tidyverse)
#library(rstanarm)
#library(rstan)
#library(tidybayes)
library(viridis)
library(cowplot)
#library(lsa)
library(kableExtra)
#rstan_options(auto_write = TRUE)
#options(mc.cores = parallel::detectCores())
#for plotting
theme_set(theme_bw())

@@

We build on work by @mankewitz2021, who created a task with the aim of bridging the gap between reference games with hard-to-name objects and negotiation datasets with easily nameable objects. They created a 3-person game where players each selected what flowers to grow from a set of 6 flowers with variable and partially hidden value. Players got the value of the flower only if they were the only one to select it, which incentivized coordination and negotiation. However, all the flowers shown were the same color, so they were not easily nameable, and the flowers repeated, leading to opportunities for ad-hoc conventionalization. This game structure encouraged players to describe flowers while never explicitly telling a player to refer to any specific flower. The experiment had two conditions, one where players within a group won points together (and thus had fully aligned incentives) and another where they won points as individuals (and didn't). @mankewitz2021 found a slight decrease in language over the course of the games, but they did not conduct a detailed analysis to assess whether the decreases in words were associated with shorter and more conventionalized referring expressions. Here, we address this gap by separating out the referring expressions in order to check for patterns of convention-formation.

@@

This paper presents a reanalysis of data from @mankewitz2021. We describe both the original data collection and the additional data processing done for reanalysis.

## Procedure
As described in @mankewitz2021, participants played a real-time coordination game in groups of three, implemented using Empirica [@almaatouqEmpiricaVirtualLab2020]. On each trial, each group saw a set of 6 flower images (Figure \ref{interface}A) . Each participant saw the values for 4 of the flowers (represented as a colored bar), such that each flower's value was hidden from one participant. Players could coordinate and discuss using a chat box before each player selected a flower. If one player selected a flower, it was worth the shown reward; if multiple players collided and selected the same flower, they each got a lower reward instead (Figure \ref{interface}B). Players needed to communicate about the flowers in order to coordinate their selections. The rewards translated into a monetary bonus for the participants at the end of the game.

@@

## Participants
@mankewitz2021 recruited `r 3 *(d.games |> filter(conditionName!="Old Cooperative") |> nrow())` participants total for 25 3-person games in each of the individual and shared utilities conditions. Some games did not complete the full 24 trials because participants dropped out early.  We excluded incomplete games from analyses, which left us with `r d.games.final |> filter(condition=="Individual Utilities") |> nrow()` games in the individual utilities condition and `r d.games.final |> filter(condition=="Shared Utilities") |> nrow()` in the shared utilities condition (`r 3*(d.games.final |> nrow())` participants included in the analyses).

@@

        top <- plot_grid(all_words, words, utts,nrow=1, labels="AUTO")
        plot_grid(top, legend, nrow=2, rel_heights=c(1,.15))

@@

@@

The classic reference game phenomenon of convergence to partner-specific referring expressions can be quantified using similarity metrics as the combination of three patterns occurring over time: 1) within a group, descriptions of the same flower converge, 2) within a group, descriptions of different flowers diverge, and 3) between groups, descriptions of the same flower diverge. We assess all of these in two ways: 1) comparing referring expressions within the game to other referring expressions from the same time to get at the dynamic changes, and 2) comparing referring expressions within the game to the names players gave the flowers after the game, to track how expressions evolve towards these final conventions.

@@

knitr::kable(text_examples, caption = "Examples of descriptions of different flowers from different games, illustrating reduction and convergence phenomenon. Games are numbered, letters refer to players within games. Images of flowers can be found in Figure 1: flower 1 upper left, flower 2 lower center, flower 3 lower left.\\label{extable}",  format="latex", booktabs=TRUE, linesep="") %>%
kable_styling(full_width = F) %>%
  row_spec(c(3,9), hline_after=T) |> 
column_spec(4, width = "12em") %>%
  kable_styling(latex_options="H")

@@

@@

```{r withinend, fig.env="figure", fig.pos = "h", fig.align = "center", out.width="90%", fig.width=4, fig.height=2, fig.cap = "Cosine similarities between flower descriptions provided by participants after the game. Error bars are 95\\% CIs, but some are hidden by the point size. \\label{withinend}"}
in_end|> 
  filter(player2!=player1) |> 
  mutate(distance=case_when(
    game1==game2 ~ "Same group",
    T ~ "Different group"),
    same_flower=ifelse(ref1==ref2, "Same flower", "Different flower"))|> 
ggplot( aes(x=str_c(same_flower,"\n", distance),y=sim,color=condition))+
  labs(y="Cosine Similarity", x=NULL)+
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width=.3))+theme(legend.position=c(0.2,.75), legend.title=element_blank(), legend.background=element_blank())+
        ggthemes::scale_color_solarized()
```
## Individual and shared utility conditions

@@

In conclusion, our analysis here showed that the phenomenon of reduction and conventionalization occurs even in complex games with different incentives, asymmetric knowledge, and more open-ended negotiation and dialogue than that found in reference games. Additionally, we saw hints that different incentives may lightly moderate the formation of conventions, but that reduction and convention formation took place regardless of incentive condition. We hope future work will continue to develop our understanding of these phenomena in more natural settings.

>>>>

output: cogsci2016::cogsci_paper
final-submission: \cogscifinalcopy
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=3, fig.height=3, fig.crop = F, 
                      fig.pos = "tb", fig.path='figs/',
                      echo=F, warning=F, cache=T, 
                      message=F, sanitize = T)
library(here)
library(tidyverse)
#library(rstanarm)
#library(rstan)
#library(tidybayes)
library(viridis)
library(cowplot)
#library(lsa)
library(kableExtra)
#rstan_options(auto_write = TRUE)
#options(mc.cores = parallel::detectCores())
#for plotting
theme_set(theme_bw())

@@

We build on work by @mankewitz2021, who created a task with the aim of bridging the gap between reference games with hard-to-name objects and negotiation datasets with easily nameable objects. They created a 3-person game where players each selected what flowers to grow from a set of 6 flowers with variable and partially hidden value. Players got the value of the flower only if they were the only one to select it, which incentivized coordination and negotiation. However, all the flowers shown were the same color, so they were not easily nameable, and the flowers repeated, leading to opportunities for ad-hoc conventionalization. This game structure encouraged players to describe flowers while never explicitly telling a player to refer to any specific flower. The experiment had two conditions, one where players within a group won points together (and thus had fully aligned incentives) and another where they won points as individuals (and didn't). @mankewitz2021 found a slight decrease in language over the course of the games, but they did not conduct a detailed analysis to assess whether the decreases in words were associated with shorter and more conventionalized referring expressions. However, this pattern of decreased language production is actually totally consistent with increased task familiarity, fatigue, and/or less communication about overall strategy, with no change in the actual referring language.  Here, we address this gap by separating out the referring expressions in order to check for patterns of convention-formation.

@@

This paper presents a reanalysis of data from @mankewitz2021. We describe both the original data collection and the additional data processing done for reanalysis. The materials, data, and code are all available at https://github.com/vboyce/AA-flowers.

## Procedure
As described in @mankewitz2021, participants played a real-time coordination game in groups of three, implemented using Empirica [@almaatouqEmpiricaVirtualLab2020]. On each trial, each group saw a set of 6 flower images (Figure \ref{interface}A) . Each participant saw the values for 4 of the flowers (represented as a colored bar), such that each flower's value was hidden from one participant. Players could coordinate and discuss using a chat box before each player selected a flower. There were no restrictions on what participants could say in the chat.

If one player selected a flower, it was worth the shown reward; if multiple players collided and selected the same flower, they each got a lower reward instead (Figure \ref{interface}B). Players needed to communicate about the flowers in order to coordinate their selections. The rewards translated into a monetary bonus for the participants at the end of the game.

@@

## Participants
@mankewitz2021 recruited `r 3 *(d.games |> filter(conditionName!="Old Cooperative") |> nrow())` participants total for 25 3-person games in each of the individual and shared utilities conditions. Games took roughly 20 minutes with wide variability (5-40 minutes). If a participant disconnected, the game ended, so some games did not complete the full 24 trials as participants dropped out early.  We excluded incomplete games from analyses, which left us with `r d.games.final |> filter(condition=="Individual Utilities") |> nrow()` games in the individual utilities condition and `r d.games.final |> filter(condition=="Shared Utilities") |> nrow()` in the shared utilities condition (`r 3*(d.games.final |> nrow())` participants included in the analyses).

@@

        top <- plot_grid(all_words, words, utts,nrow=1, labels="AUTO", hjust=c(-5, -4, -4), vjust=1.8)
        plot_grid(top, legend, nrow=2, rel_heights=c(1,.15))

@@

Participants employed a range of reference strategies, including referring to the flowers by a common name (ex. rose, iris), referring to the number or groupings of flowers, referring to properties of the flowers (ex. spiky), referring to background elements (ex. white house), and analogies to other objects (ex. flame). Descriptions sometimes combined multiple of these approaches.

@@

The classic reference game phenomena of convergence to partner-specific referring expressions can be quantified using similarity metrics as the combination of three patterns occurring over time: 1) within a group, descriptions of the same flower converge, 2) within a group, descriptions of different flowers diverge, and 3) between groups, descriptions of the same flower diverge. We assess all of these in two ways: 1) comparing referring expressions within the game to other referring expressions from the same time to get at the dynamic changes, and 2) comparing referring expressions within the game to the names players gave the flowers after the game, to track how expressions evolve towards these final conventions.

@@

knitr::kable(text_examples, caption = "Examples of descriptions of different flowers from different games, illustrating reduction and convergence phenomena. Games are numbered, letters refer to players within games. Images of flowers can be found in Figure 1: flower 1 upper left, flower 2 lower center, flower 3 lower left.\\label{extable}",  format="latex", booktabs=TRUE, linesep="") %>%
kable_styling(full_width = F) %>%
  row_spec(c(3,9), hline_after=T) |> 
column_spec(4, width = "12em") %>%
  kable_styling(latex_options="H")

@@

```{r withinend, fig.env="figure", fig.pos = "h", fig.align = "center", out.width="90%", fig.width=4, fig.height=2, fig.cap = "Cosine similarities between flower descriptions provided by participants after the game. Error bars are 95\\% CIs, but some are hidden by the point size. \\label{withinend}"}
in_end|> 
  filter(player2!=player1) |> 
  mutate(distance=case_when(
    game1==game2 ~ "Same group",
    T ~ "Different group"),
    same_flower=ifelse(ref1==ref2, "Same flower", "Different flower"))|> 
ggplot( aes(x=str_c(same_flower,"\n", distance),y=sim,color=condition))+
  labs(y="Cosine Similarity", x=NULL)+
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width=.3))+theme(legend.position=c(0.2,.75), legend.title=element_blank(), legend.background=element_blank())+
        ggthemes::scale_color_solarized()
```

@@

## Individual and shared utility conditions

@@

In conclusion, our analysis here showed that the phenomena of reduction and conventionalization occurs even in complex games with different incentives, asymmetric knowledge, and more open-ended negotiation and dialogue than that found in reference games. Additionally, we saw hints that different incentives may lightly moderate the formation of conventions, but that reduction and convention formation took place regardless of incentive condition. We hope future work will continue to develop our understanding of these phenomena in more natural settings.
